Balsa: A Compact Line Notation Based on SMILES

### 2022-10-23 manuscript.tex
Makes line edit.
Adds Acknowledgement, fixes several minor issue.

Compatibility between Balsa and ProtoSMILES is summarized in Table~\ref{table:feature-comparison}. Barring elements of ProtoSMILES that are either underspecified or internally inconsistent (entries marked \enquote{maybe}), Balsa can be seen to be a subset of ProtoSMILES. Output from Balsa writers should therefore be readable by SMILES readers. However, output from SMILES writers may or may not be readable by Balsa readers.

@@

Another advantage is lossless interconversion with other serialization formats. The line notation InChI might appear to compete with Balsa in this sense. The authors of InChI have noted that InChI is not a solution to the molecular serialization problem but rather an identifier.\cite{heller:2015} The reason presumably lies with the fact that InChI uses concepts such as hydrogen delocalization that are uncommon elsewhere. To date, no third-party reader or writer of InChI has been reported.

@@

This paper describes Balsa, a compact molecular serialization format designed for backward-compatibility with SMILES. Constitutionally, Balsa can encode any molecule conforming to the valence bond model. In the event of undesired symmetry artifacts due to delocalization, Balsa offers a mitigation based in graph theory. Conformational isomerism of alkenes is supported by partial parity bonds. The configurations of tetracoordinate, stereogenic atoms are encoded through the use of a parity enumeration and conventions around its use.

@@

The complete set of rules needed to interpret Balsa's syntax are described in detail. This is a crucial component of the language's definition because Balsa achieves its information density largely by favoring convention over encoding. As an aid to difficult cases, guidelines for readers and writers are included.

@@

>>>>

Compatibility between Balsa and ProtoSMILES is summarized in Table~\ref{table:feature-comparison}. Barring elements of ProtoSMILES that are either underspecified or internally inconsistent (entries marked \enquote{maybe}), Balsa can be seen to be a subset of ProtoSMILES. Output from Balsa writers should therefore be readable by SMILES readers. However, output from SMILES writers may or may not be readable by Balsa readers. Balsa can therefore be characterized as forward compatible with ProtoSMILES~\cite{extendingVersioningLanguages}.

@@

Another advantage is lossless interconversion with other serialization formats. The line notation InChI might appear to compete with Balsa in this sense. The authors of InChI have noted that InChI is not a solution to the molecular serialization problem but rather an identifier~\cite{heller:2015}. The reason presumably lies with the fact that InChI uses concepts such as hydrogen delocalization that are uncommon elsewhere. To date, no third-party reader or writer of InChI has been reported.

@@

This paper describes Balsa, a compact molecular serialization format designed to be forward compatible with SMILES. Constitutionally, Balsa can encode any molecule conforming to the valence bond model. In the event of undesired symmetry artifacts due to delocalization, Balsa offers a mitigation based in graph theory. Conformational isomerism of alkenes is supported by partial parity bonds. The configurations of tetracoordinate, stereogenic atoms are encoded through the use of a parity enumeration and conventions around its use.

@@

The complete set of rules needed to interpret Balsa's syntax are described in detail. This is a crucial component of the language's definition because Balsa achieves its information density in part by favoring convention over encoding. As an aid to difficult cases, guidelines for readers and writers are included.

@@

\section*{Acknowledgement}

I thank Andrew Dalke and John Mayfield for helpful conversations during the preparation of this manuscript.
